---
title: "20250206_autohsct_MEPHISTO"
author: "jenssle"
date: "06 02 2025"
output: html_document
---

## Setup 

Load the packages 

```{r}
library(MOFA2)
library(phyloseq)
library(tidyverse)
library(microViz)
library(readxl)

`%not_in%` <- purrr::negate(`%in%`)
```

Load the processed 16S data in form of a phyloseq object.

```{r}
ps <- readRDS("/home/FFM16S_phylo.rds")
```

## 16S processing

Filter for OTUs that only appear in at least 1 percent of our samples and transform using robust centered log ratio transformation (rclr).

```{r}
ps2 <- ps %>%
  tax_filter(min_prevalence = 0.005, 
             min_total_abundance = 10) %>%
  tax_fix() %>% 
  tax_agg(rank = "Family")

#Subset to those taxa identified on the family level
ps3 <- subset_taxa(ps2, Family %in% colnames(otu_table(ps2) %>%
                                               as.data.frame())[!grepl(
                                                 "roup|class|Class|Kingdom|TC|Order|Phylum|Mitochondria|UCG|vadin",
                                                 colnames(otu_table(ps2) %>%
                                                            as.data.frame())
                                               )])
#Now, perform the transformance
ps3 <- tax_transform(ps3, "rclr")


df_otu <- otu_table(ps3) %>% 
  as.data.frame() 
```

Read in the timepoint info for the timepoint-acounting modelling. 

```{r}
meta_df <- read_excel("/home/FFM16S_metadata.xlsx") %>%
  dplyr::select(Unique_ID, Pat_ID, Timepoint, Diffdays, InvSimpson, Mibi_ID) %>%
  mutate(Mibi_ID = as.character(Mibi_ID))
```

Extract the OTU table and shape for the use in MEFISTO.

```{r}
library(psych)
library(readxl)

#bring it into long shape
df_otu_long <- df_otu %>%
  rownames_to_column("seq_ID") %>%
  pivot_longer(!seq_ID, names_to = "feature", values_to = "value")

#add the meta information which is stored in the ps objects meta data
df_meta <- sample_data(ps) %>%
  as.data.frame()

df_meta$seq_ID <- rownames(sample_data(ps))

df_meta <- left_join(
  as.tibble(df_meta) %>%
    dplyr::select(-Timepoint),
  meta_df,
  by = "Mibi_ID"
) %>%
  distinct(internal_n, .keep_all = TRUE) %>%
  drop_na(Diffdays) %>%
  mutate(Mibi_ID = as.character(Mibi_ID)) %>%
  mutate(ent = case_when(grepl("Myelom|Plasmozytom", entity) ~ "MM",
                         TRUE~ "other"))

df_otu_long <- left_join(
  df_otu_long %>%
    dplyr::filter(seq_ID %in% df_meta$seq_ID), 
  df_meta[,c("seq_ID", "Mibi_ID", "ID", "Diffdays", "entity", "InvSimpson")], 
  by = "seq_ID"
) 

df_mofa <- df_otu_long %>%
  dplyr::rename(
    c("group" = "ID"), 
    c("sample" = "seq_ID")
  ) %>%
  filter(group != "Mock") %>%
  mutate(view = "16S") %>%
  drop_na(Diffdays)
```

## Process immune data

Read in the immune data and bring to right format.

```{r}
df_immuno <- read_excel("/home/FFMImmuneData.xlsx")

#collect the immune cell names
immun_cells <- colnames(df_immuno)[22:51]

df_immuno1 <- df_immuno %>%
  dplyr::select(c("Mibi_ID", immun_cells)) %>%
  mutate(Mibi_ID = as.character(Mibi_ID)) %>%
  pivot_longer(!Mibi_ID, names_to = "feature", values_to = "value") %>%
  left_join(
        df_mofa %>%
          filter(Mibi_ID %in% df_immuno$Mibi_ID) %>%
          dplyr::select(
            "sample",
            "Mibi_ID",
            "group",
            "Diffdays",
            "entity",
            "view",
            "InvSimpson"
          ) %>%
          mutate(view = "immuno") %>%
          distinct(sample, .keep_all = TRUE),
        by = "Mibi_ID"
      ) %>% drop_na(sample)


df_mofa_immuno <- df_mofa %>%
  rbind(df_immuno1)

df_mofa_immu <- df_mofa %>%
  filter(sample %in% df_immuno$sample) %>%
  rbind(
    df_immuno %>%
      filter(sample %in% df_mofa$sample) %>%
      left_join(
        df_mofa %>%
          filter(sample %in% df_immuno$sample) %>%
          dplyr::select(
            "sample",
            "Mibi_ID",
            "group",
            "Diffdays",
            "entity",
            "view",
            "InvSimpson"
          ) %>%
          mutate(view = "immuno") %>%
          distinct(sample, .keep_all = TRUE),
        by = "sample"
      )
  )
```

## Prepare and train MOFA

Create MOFA object

```{r}
mofa_ob_untr <- create_mofa(data = df_mofa %>%
                              mutate(group = "1"))

mofa_ob_untr
```

Set the timevariable.

```{r}
mofa_ob_untr <- set_covariates(mofa_ob_untr,
                                       covariates = "Diffdays")
```

Show the data overview.

```{r}
plot_data_overview(mofa_ob_untr,
                               show_covariate = FALSE,
                               show_dimensions = FALSE) 
```

Set model options.

```{r}
data_opts <- get_default_data_options(mofa_ob_untr)
data_opts$center_groups <- FALSE

model_opts <- get_default_model_options(mofa_ob_untr)
model_opts$num_factors <- 15

mefisto_opts <- get_default_mefisto_options(mofa_ob_untr)
mefisto_opts$n_grid <- 10
mefisto_opts$start_opt <- 50
mefisto_opts$opt_freq <- 50
mefisto_opts$model_groups <- FALSE 

train_opts <- get_default_training_options(mofa_ob_untr)
train_opts$seed <- 069
```

Prepare model.

```{r}
mofa_ob_untr <- prepare_mofa(
  object = mofa_ob_untr,
  data_options = data_opts,
  model_options = model_opts,
  training_options = train_opts,
  mefisto_options = mefisto_opts
) 
```

Train the MEFISTO model. 

```{r}
##################################
## Define python version to use ##
##################################

reticulate::use_python("/home/jenssle/miniconda3/bin/python", required = TRUE)

#####################
## Train the model ##
#####################

outfile <- paste0(getwd(),"/home/_mefisto_16S_imuno")
mofa_ob_tr <- run_mofa(mofa_ob_untr,
                       outfile,
                       use_basilisk = TRUE
                       )
```

## Collect data

Extract the factors and save.

```{r}
mefisto_latent <- get_factors(mofa_ob_tr, factors = "all")$`1` %>%
  as.data.frame() 

saveRDS(mefisto_latent, file = "/home/mefisto_LF.rds")
```
