---
title: "dada2 based miseqrun processing"
author: "ensslejc"
format: 
  gfm : 
    code-fold: false
editor: visual
date: "`r Sys.Date()`"
knitr:
  opts_chunk:
    message: false
    warning: false
---

This is the standard script to process 16S sequencing data from fecal samples. Per miseq run, a mock sample was included to determine the accuracy of the taxonomy inference.

## Setup

### Packages

```{r, eval = FALSE}
library(tidyverse)
library(dada2)
```

### unzip files

Unzip your fastq.

```{bash, eval = FALSE}
cd /home/your_dir/MISeq_run/unzipped

gzip -d *.gz
```

### Set path

```{r, eval = FALSE}
path <- "~/your_dir/MISeq_run/unzipped"

list.files(path= ath)
```

## Processing

### Read in the names

```{r, eval = FALSE}
# Forward and reverse fastq filenames have format: SAMPLENAME_R1_001.fastq and SAMPLENAME_R2_001.fastq
fnFs <- sort(list.files(path, pattern="_R1_001.fastq", full.names = TRUE))
fnRs <- sort(list.files(path, pattern="_R2_001.fastq", full.names = TRUE))
# Extract sample names, assuming filenames have format: SAMPLENAME_XXX.fastq
sample.names <- sapply(strsplit(basename(fnFs), "_L001"), `[`, 1)
```

### Inspect the quality profiles

```{r, eval = FALSE}
plotQualityProfile(fnFs[1:6])
```

### Filter and trim

Assign the filenames for the filtered fastq.gz files.

```{r, eval = FALSE}
# Place filtered files in filtered/ subdirectory
filtFs <- file.path(path, "filtered", paste0(sample.names, "_F_filt.fastq.gz"))
filtRs <- file.path(path, "filtered", paste0(sample.names, "_R_filt.fastq.gz"))
names(filtFs) <- sample.names
names(filtRs) <- sample.names
```

Run the filtering with standard parameters. Trim and truncation was optimized by manual inspection of grid search.

```{r, eval = FALSE}
out <-
  filterAndTrim(
    fnFs,
    filtFs,
    fnRs,
    filtRs,
    trimLeft = c(17,21),
    truncLen = c(285, 240), 
    maxN = 0,
    maxEE = c(2, 2),
    truncQ = 2,
    rm.phix = TRUE,
    compress = TRUE,
    multithread = as.integer(60)
  ) 

head(out)
```

### Learn the error rates

```{r, eval = FALSE}
errF <- learnErrors(filtFs, multithread=as.integer(40))

errR <- learnErrors(filtRs, multithread=as.integer(40))
```

Make sure that zero read files are removed from the sample name list.

```{r, eval = FALSE}
filtFs <- filtFs[file.exists(filtFs)]
filtRs <- filtRs[file.exists(filtRs)]
```

### Sample inference

```{r, eval = FALSE}
dadaFs <- dada(filtFs, err=errF, multithread=as.integer(40))

dadaRs <- dada(filtRs, err=errR, multithread=as.integer(40))
```

Check the returned dada class object.

```{r, eval = FALSE}
dadaFs[[1]]
```

### Merge paired reads

```{r, eval = FALSE}
mergers <- mergePairs(dadaFs, filtFs, dadaRs, filtRs, verbose=TRUE)
# Inspect the merger data.frame from the first sample
head(mergers[[1]])
```

### Construct sequence table

```{r, eval = FALSE}
seqtab <- makeSequenceTable(mergers)
dim(seqtab)
```

### Inspect the distribution of sequence lengths

```{r, eval = FALSE}
# Inspect distribution of sequence lengths
table(nchar(getSequences(seqtab)))
```

### Remove chimeras

```{r, eval = FALSE}
seqtab.nochim <- removeBimeraDenovo(seqtab, method="consensus", multithread=as.integer(40), verbose=TRUE)
dim(seqtab.nochim)

sum(seqtab.nochim)/sum(seqtab)
```

### Track reads through the pipeline

```{r, eval = FALSE}
getN <- function(x) sum(getUniques(x))
track <- cbind(out, sapply(dadaFs, getN), sapply(dadaRs, getN), sapply(mergers, getN), rowSums(seqtab.nochim))

colnames(track) <- c("input", "filtered", "denoisedF", "denoisedR", "merged", "nonchim")

#add the fraction information
track <- track %>% 
  as.data.frame() %>%
  mutate(perc_input = nonchim/input, 
         perc_filtered = nonchim/filtered, 
         perc_merged = nonchim/merged)

rownames(track) <- sample.names
head(track)
```

Visualize the respective readouts.

```{r, eval = FALSE}
library(cowplot)

plot_grid(
  ggplot(track, aes(nonchim)) +
    geom_histogram() +
    theme_bw(),
  
  ggplot(track, aes(perc_input)) +
    geom_histogram() +
    theme_bw()
)
```

### Assign taxonomy

To do so, the DADA2 implementation of the naive Bayesian classifier method is applied.

```{r, eval = FALSE}
taxa <-
  assignTaxonomy(seqtab.nochim,
                  "/home/your_dir/silva_nr99_v138.1_train_set.fa.gz",
                 multithread=as.integer(70))
```

Lets investigate the assignment.

```{r, eval = FALSE}
taxa.print <- taxa 
rownames(taxa.print) <- NULL
head(taxa.print, 50)
```

## Phyloseq workup

```{r, eval = FALSE}
library(phyloseq)
library(Biostrings)
library(ggplot2)
```

Read in the meta data.

```{r, eval = FALSE}
metadf <- read.delim("/home/your_dir/MISeq_run/sample_meta.txt", header=FALSE, comment.char="#")

colnames(metadf) <- c("sample", "internal_n", "sample_id", "part_of_study", "Index_1", "Index_2")

rownames(metadf) <- metadf$sample

seqtab.nochim2 <- seqtab.nochim %>%
  as.data.frame() %>%
  rownames_to_column("rwn") %>%
  separate(rwn, c("keep", "remove"), sep = "_S")

rownames(seqtab.nochim) <- seqtab.nochim2$keep

metadf1 <- metadf[match(rownames(seqtab.nochim), metadf$internal_n),] %>%
  mutate(internal_n = as.character(internal_n)) %>%
  left_join(track %>% 
              rownames_to_column("rwn") %>%
              separate(rwn, c("internal_n", "remove"), sep = "_S") %>%
              mutate(internal_n = as.character(internal_n)) %>%
              dplyr::select(internal_n, perc_input), 
            by = "internal_n")%>%
  drop_na(internal_n)

rownames(metadf1) <- metadf1$internal_n

study_samples <- (metadf1 %>% filter(perc_input > 0.4, 
                                     part_of_study == "yes"))$sample
```

Construct a phyloseq object from the dada2 output.

```{r, eval = FALSE}
ps <- phyloseq(
  otu_table(seqtab.nochim, taxa_are_rows = FALSE),
  sample_data(metadf1),
  tax_table(taxa)
)
```

Investigate the mock sample as quality control.

```{r, eval = FALSE}
ps.mock <- ps
ps.mock <- prune_samples(sample_names(ps.mock) == "20210715-95", ps.mock) 
```

Visualize via barplot. The theoretical composition should be as follows: Listeria monocytogenes - 12%, Pseudomonas aeruginosa - 12%, Bacillus subtilis - 12%, Escherichia coli - 12%, Salmonella enterica - 12%, Lactobacillus fermentum - 12%, Enterococcus faecalis - 12%, Staphylococcus aureus - 12%, Saccharomyces cerevisiae - 2%, and Cryptococcus neoformans - 2%.

```{r, eval = FALSE}
top20 <- names(sort(taxa_sums(ps.mock), decreasing=TRUE))[1:20]
ps.top20 <- transform_sample_counts(ps.mock, function(OTU) OTU/sum(OTU))
ps.top20 <- prune_taxa(top20, ps.top20)
plot_bar(ps.top20, x="internal_n", fill="Genus")
```

Now, subset for samples part of the study:

```{r, eval = FALSE}
ps_study <- prune_samples(sample_names(ps) %in% rownames(metadf1 %>% filter(perc_input > 0.4, 
                                     part_of_study == "yes")), ps) 
```

Save the phyloseq object.

```{r, eval = FALSE}
saveRDS(ps_study, "/home/your_dir/miseqrun_phyloseq_processed.rds")
```
